Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-14 draft.tex
First correction

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous data variables describing the medical condition as well as demographic and lifestyle factors of a subject.
%
Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations ('\emph{interactions}').
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
\\
We propose a Interactive Visual Analytics approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
%We propose a Interactive Visual Analytics approach to provide epidemiologists with the tools necessary to analyze this data.
%
It allows for both classical hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
%
Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using hierarchical agglomerative clustering. (Remove Clustering reference?)
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims to characterize health and disease by determining risk factors.
%
Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
%
%Cohort studies are a epidemiological method to gather data about a group of subjects (a \emph{cohort)}) to make general statements about these problems.
%
Observations made by clinicians in the daily routine are translated into hypothesis.
%
These are used to determine environmental and lifestyle factors are as well as medical attributes which are believed to influence a condition of interest.
%
The data variables necessary are gathered using interviews and clinical examinations.
%
Statistical methods like regression analysis aim aim to check the attribute list for plausibility.
%

Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection bias prohibit statements based on statistical correlations in the cohort to be inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
%
Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
Statistical correlations are also prone to confounding, meaning that two factors influence each other and therefore should be normalized with respect to each other.
%
When for example one investigates risk factors for prostate cancer in male subjects the outcome is strongly dependent on the age.
%
Therefore results need to be age adjusted to be comparable.
%
Confounding variables, however, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data which introduces new problems.
%
Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to segment it.
%
Manual segmentation via radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with other epidemiological factors require techniques which reach beyond standard statistical methods.
\\
We propose a Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
%
\textbf{ToDo Characterization of the healthy aging process! Which change indicates a unusual pathological change?}

%Interactive Visual Analysis (IVA) TODO QUELLE provides techniques which not only show promising potential for combining both 3d visualization with

Our contributions are:
\begin{itemize}
	\item Applying the Interactive Visual Analysis approach to the epidemiological problem domain by characterizing special affordances of this context.
	\item Provide an overview over the workflow for analyzing cohort study data to gain insigth into the large subject spaces.
	\item Provide Visualization Techniques which combine both Information Visualization and 3D Rendering of Organ Shapes as well as combining them with well known epidemiological graphics and key figures.
	\item Implement the presented methods in a Web Framework based on WebGL, D3js and Nodejs as backend.
\end{itemize}

@@

\section{Medical and Technical Background} \label{MedicalAndTechnicalBackground}
Wer ist an epidemiologischen Studien beteiligt?\\
•  Ärzte (Facharzt für öffentliches Gesundheitswesen, Gene@ker)\\
•  Medizinische Informa@ker mit Fokus auf Biometrie und Sta@s@k\\
•  Bei klinischen Studien: Ärzte des entsprechenden Fachs

In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
ToDo Define Epidemiological Outcome
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis.
%
Following Thew and colleagues publication on this matter, the workflow can be characterized as follows.
%
Hypothesis most commonly base on observations made by clinicians in their daily routine.
%
A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
%
Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
%
Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
%

Reproducibility of results is an epidemiological key requirement.
%
Longitudinal epidemiological studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
%
This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, relative risks a expressed through the evaluation of p values which indicate statistical significance.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiolgical data. 
%
Graphic data representation is largely used to show results rather than gaining insight.
\\
Group subjects using epidemiological factors is essential in order to make statements about their statitical power.
%
Grouping is carried out hypothesis driven.
%
Age for example is also divided into groups (\emph{binned}) when investigating its influence on a condition.
%
These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data is highly heterogenous.
%
Information about medical history and examinations, genetic conditions, geographical data, questionaire answers, and image data yield complex data spaces for each subject.
%
Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
%
This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
%
Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
%

@@

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
%
MRI data is more expensive to obtain as CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
%
The quality of medical image data acquired for cohort studies is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.

\paragraph{Image analysis.} \label{ImageAnalysis} When analyzing medical image data there have decisions to be made on how they are \emph{compared} and \emph{quantified}.
%
Segmentations masks describing all parts of the shape of interest would be ideal since many different key figures can be derived from them.
%
Since these masks require sophisticated algorithms custom tailored to the data sets the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks are also not nearly as expressive and versatile as segmentation masks.
%
They are also prone to a high inter-observer variability and hard to reproduce.
%
This gains even more momentum when analyzing multiple time steps!
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determine type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected independent of diseases in mind.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
%
The MRI protocol features a rich number of different sequences.
%
Also for women, breast MRI scans were acquired.
%
A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
%
The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

Finding a visualization which fits all needs and communicates all aspects of the data equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}
%
Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but making the influence of each variable hard to determine.
\\\\
The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process when analyzing the data.
%
Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
%
These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
%
This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
This of course requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
The technique was applied to variables generated by segmenting the the brain into 45 parts and measure the number of voxel, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive derived measures of the epidemiological factors.
%
Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights certain statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

Gresh and colleagues suggest with \texttt{WEAVE} one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
%
This approach already incorporated Data Mining approaches like dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
%
Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views like parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiolgical context.

Commercial systems like Tableau or Spotfire provide a rich user interface that allows to apply Visual Analytics techniques without the need of writing any code.
%
With little efford linked views can be created using these tools, but the data procesinng possibilities like derivation of new variables or the volume rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language

% *********************** Our own Work ***********************
Klemm and colleagues used captured lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering was used to divide the population into shape-related groups.
%
As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured and behaved as expected.
%
This work focuses on incorporating these data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques on the non-image data is was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
% *********************** / Our own Work ***********************

Generalized Pairs Plot (\emph{GPLOMS}) is an information visualization technique that allows for heterogenous variables to be pairwise visualized using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
%
This technique is also useful to gain an overview over numerous variables and their distributions.
%
It uses histograms, bar charts, scatterplots and heat maps to visualize the different variable combinations with regard to their type.
%
Brushing capabilities allow for brushing and linking as well as filtering, but has limitations like making only one category brushable at a time.
%
We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variable but does not fit in the scope of this paper.
%
A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots integrated regression lines \cite{Dai2005}.
%
From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.

@@

\section{Interactive Visual Analytics in Cohort Study Data}
\begin{figure}[htb]
 \centering
 \label{fig:WorkflowComparison}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{Visual Analytics systems are able to complement parts of the epidemiological workflow, not to replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
\end{figure}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Variable of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types. The data is visualized in the independent and dependent domain space, which can then be brushed, yielding new groups which can be investigated using further variables. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
\end{figure}

As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a strict sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
%
Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as steps consecutive series of steps.
%
The workflow we propose by introducing the \emph{IVA} principle into the epidemiological problem domain does not aim to replace the existing workflow but to complement its weaknesses 
%
In the current state the workflow handles the data as a black box.
%
A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
The resulting value decides whether the data supports the hypothesis or not.
%
It would be possible that there actually are features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach it is not able to find those.
%
This becomes even more imminent when the workflow is adapted to the analysis of the medical image data.
%
Domain experts annotate tediously landmarks which allow to derive metrics like distances which are then handled like other features and analyzed using the same set of statistical tools.
%
Not only does this leave out the majority of the information in the medical image data by abstracting it to metrics, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%
Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
IVA tries to illuminate the data black box by making the domain experts part of the feature list selection process.
%
Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
Note that the process also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.

According to Steffens Terminology in Interactive Visual Analysis of Perfusion Data
\\\\

>>>>

%% Abstract section.
\abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
Analyzing these data requires image segmentation, extraction of key figures and shape based subject grouping.
\\\\
We propose a Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as sociodemographic and medical attribute data.
%
It allows for both hypothesis validation approaches as well as hypothesis generation by incorporating data mining methods.
%
Adaptive linked information visualization views and 3d-shape renderings are combined with epidemiological techniques.
%
Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims to characterize health and disease by determining risk factors.
%
Clinical problems and questions answered using epidemiological methods comprise diagnosis accuracy, disease frequency, risk factors, disease prognosis, effectiveness of treatments or preventions and cause of diseases \cite{Fletcher2012}.
%
%Cohort studies are a epidemiological method to gather data about a group of subjects (a \emph{cohort)}) to make general statements about these problems.
%
Observations made by clinicians in the daily routine are translated into hypothesis.
%
These are used to determine environmental and lifestyle factors are as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Statistical methods like regression analysis aim aim to check the attribute list for plausibility.
%

Longitudinal population-based studies like the Study of Health in Pomerania \cite{Volzke2011} aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The sample is drawn randomized to avoid selection bias which prohibits statements based on statistical correlations in the cohort from being inferred to the whole population (\emph{external validity}) \cite{Fletcher2012}.
%
Also a information bias needs to be avoided by strictly standardizing the data acquisition.
%
Statistical correlations are also prone to \emph{confounding}, meaning that two factors are dependent and therefore should be normalized with respect to each other.
%
When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
%
Therefore results need to be age adjusted to be comparable.
%
Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
\\\\
Modern cohort studies often include medical image data which introduces new problems.
%
Since it is unethical to expose people to radiation, non-harming imaging like Magnetic Resonance Imaging (MRI) or Ultrasound Imaging is used.
%
As MRI scans are expensive there exists a tradeoff between quality of the image data and their associated costs.
%
To quantify these data it is necessary to label each voxel regarding structure affiliation (\emph{segmentation}).
%
Manual segmentation carried out by radiological experts is possible but very costly and prone to inter- and intra observer variability.
%
Segmentation algorithms allow for (semi)-automated analysis of the data but require sophisticated methods due to high inter-subject variability caused by the subject diversity.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.
\\\\
We propose a Interactive Visual Analysis approach \cite{Thomas2005} to provide a way to analyze both image- and non-image data.
%
Visual queries and direct feedback of Visual Analytics systems allow for a fast exploration of the data space.
%
Intended as a extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger hypothesis generation using Data Mining methods such as clustering.
%
In order to characterize the healthy aging process we aim to determine changes for subjects which indicate unusual pathological changes.
\\\\
Our contributions are:
\begin{itemize}
	\item Applying the Interactive Visual Analysis technique set to the epidemiological problem domain by characterizing affordances of this context.
	\item Provide an overview over the workflow for analyzing cohort study data to gain insight into the large diverse subject spaces.
	\item Provide visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.
	\item Implement the presented methods as a web framework based on WebGL, D3JS and NodeJS.
\end{itemize}

@@

\section{Medical and Technical Background} \label{MedicalAndTechnicalBackground}
% Wer ist an epidemiologischen Studien beteiligt?\\
% •  Ärzte (Facharzt für öffentliches Gesundheitswesen, Gene@ker)\\
% •  Medizinische Informa@ker mit Fokus auf Biometrie und Sta@s@k\\
% •  Bei klinischen Studien: Ärzte des entsprechenden Fachs

In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
% TODO Define Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
% TODO Define who participates in epidemiological studies ?
Since it is a diverse science, many different experts work at epidemiological studies, ranging from specialized doctors, medical computer scientists with focus on biometrics to statisticians.
%
Epidemiologists follow a strict workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
%
Following Thew and colleagues publication on this matter, the workflow can be characterized as follows.
%
\begin{itemize}
	\item Hypothesis most commonly base on observations made by clinicians in their daily routine.
%
	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding variables need to be adjusted so that they do not affect the effect size of a attribute.
%
	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{itemize}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).

Reproducibility of results is an key requirement.
%
Longitudinal studies require the acquired attributes to be comparable to evaluate them.
%
If the data acquisition process changes, a information bias is introduced to the data, disallowing inference between acquisition cycles.
%
This underlines the high quality standards to methods processing the data, whether to extract additional parameters or gain insight.
%
To determine, whether a subject is prone to be affected by a certain disease, relative risks are expressed through the evaluation of p-values which indicate statistical significance.
%
Statistics tools such as SPSS and STATA play a major role for analyzing epidemiological data. 
%
Graphic data representation is largely used to present results rather than gaining insight.

Grouping subjects using epidemiological factors is essential in order to allow per-group risk determination.
%
Grouping is carried out hypothesis driven.
%
Age for example is also divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
%
These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data is highly heterogenous.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yields a complex data space for each subject.
%
Often data are derived from acquired variables to either group or threshold values or get information derived from reviewed data such as breast density data for women.
%
This underlines also the problem of missing data since for ethical, legal or medical reasons some data variables can not be gathered for each subject.
%
Follow-up examinations or -questions for conditions also produce variables only available for a small amount of subjects.
%

@@

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting an hazardous amount of radiation for the subject are not suited for ethical reasons.
%
MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentations masks labeling the voxel of a shape of interest would be ideal since many different key figures can be derived from them.
%
Since these masks require sophisticated algorithms custom tailored to the data sets the epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks are also not nearly as expressive and versatile as segmentation masks.
%
They are also prone to a high inter-observer variability and hard to reproduce.
%
This gains even more momentum when analyzing multiple time steps!
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determine type of tissue).

@@

\subsection{The Study of Health in Pomerania (SHIP)}
Starting 1997 with a cohort consisting of 4.308 subjects this cohort study located in northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected diseases-independent.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
%
The MRI protocol features a rich number of different sequences.
%
Also for women, breast MRI scans were acquired.
%
A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
%
The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

Designing a visualization which communicates all aspects of the data equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
%
Data mining tools like the Principal Component Analysis are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
\\\\
The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
%
Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
%
These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
%
This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.
%
Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems which analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
%
This approach already incorporated Data Mining approaches like dividing the data space using a k-nearest-neighbor technique and Principal Component Analysis.
%
Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views like parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiolgical context.

Commercial systems like \texttt{Tableau} or \texttt{Spotfire} provide a rich user interface that allow to apply Visual Analytics techniques without the need of writing any code.
%
With little effort, linked views can be created using these tools, but the data processing possibilities like derivation of new variables or the volume rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language

% *********************** Our own Work ***********************
Klemm and colleagues used lumbar spine variabilities based on an semi-automatic shape-detection algorithm of 490 participants of the \texttt{SHIP-2} \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept a relation between size of the segmented shape and measured size of the subjects was measured and behaved as expected.
%
This work focuses on incorporating these data as new features of the overall data set, making it possible to include it into the hypothesis validation and generation process.
%
When applying clustering techniques on the non-image data is was found that \texttt{k-Prototypes} and \texttt{DBSCAN} is appropriate in the epidemiological context but is strongly dependent on the chosen variables and distance measure \cite{Klemm2014BVM}.
% *********************** / Our own Work ***********************

Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique that allows for heterogenous variables to be pairwise compared using appropriate plots in a plot matrix grouped by type \cite{Francois2013}.
%
This technique is also useful to gain an overview over numerous variables and their distributions.
%
It uses histograms, bar charts, scatterplots and heat maps to visualize the different variable combinations with regard to their type.
%
Brushing capabilities allow for brushing and linking as well as filtering, but has limitations like making only one category brushable at a time.
%
We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variable but does not fit in the scope of this paper.
%
The inspiration on the chosen visualization techniques stems from this publication.
%
A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.

@@

\section{Interactive Visual Analytics in Cohort Study Data}
\begin{figure}[htb]
 \centering
 \label{fig:WorkflowComparison}
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{ToDo Abbildung noch nicht final. Visual Analytics systems are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
\end{figure}
%
As described in subsection~\ref{EpidemiologicalWorkflow}, the epidemiolgical workflow is a strict sequence of steps taken by domain experts and needs to be reproducible and comprise statistical integrity.
%
Figure~\ref{fig:WorkflowComparison} (a) describes this workflow as consecutive series of steps.
%
The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
%
In the current state the workflow treats the data like a black box.
%
A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
The resulting value decides whether the data supports the hypothesis or not.
%
It would be possible that there actually are features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
This becomes even more imminent when the workflow is adapted to the analysis of the medical image data.
%
Domain experts annotate tediously landmarks which allow to derive metrics like distances which are then handled like other features and analyzed using the same set of statistical tools.
%
Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%
Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of the feature list selection process.
%
Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care since overfitting of expectations to the available data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.

\subsection{Image Centric Cohort Study Data in Interactive Visual Analytics Context}
In the \emph{IVA} context, data is divided into two major view types.
%
The human body exposing shape information for the \emph{physical view}, also referred as the \emph{independent variable} \cite{Oeltze2013}.
%
This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
%
%Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
%
\emph{Dependent variables} in the epidemiological context can be divided twofold:
\begin{itemize}
	\item Variables derived from the image data. 
	%
	These measures abstracts shape information as quantification to allow for comparison. 
	%
	These variables describe image data and can also be used to brush in the image space.
	\item Epidemiological socio-demographic or medical attribute data.
	%
	These values belong to every subject which is represented in the image space, but does not describe shape information.
	%
	This is the data epidemiologists usually want to correlate with image data.
\end{itemize}
%Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
%
%Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
% \textbf{Object Space}\\
% - Medical Image Data / Spine Segmentations\\
% \textbf{Attribute Space}\\
% - SHIP-Variables\\
% - Derived visualizations
% 
% ? Where to incorporate Levels of IVA
